A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2021-10-04 analysis/paper/paper.Rmd
A few more sites. Remembered to scale plot height when multiphase sites.

# Introduction
The association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C. Brøgger and his son, the archaeologist A.W. Brøgger, in their respective theses both published in 1905 (see also @maccurdy1906). Ever since, this finding has been fundamental to Stone Age research in Norway. One of its major implications is that relative sea-level change can be used for the dating of the sites. This is based on the fact that since most Stone Age sites were situated on or close to the contemporary shoreline during their original occupation, a reconstruction of the trajectory of shoreline displacement can be used to date the sites based on their elevation relative to the present day sea level. The method of shoreline dating is still fundamental to most Stone Age research in Norway, and is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available. Given the generally coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating by means of shoreline displacement is often the only and most accurate method by which one can hope to date the sites. Dating by means of shoreline displacement is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because shoreline dating is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line. Consequently, adherence or deviation from this pattern has major implications for the socio-economic foundations of the societies in question.

Despite the fundamental role of shoreline-dating in Norway, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age. This can be used both to inform the certainty with which one can depend on dating by means of shoreline displacement, it can potentially identify societal shifts if there is chronological variation in the degree to which the sites are shore-bound, and any such variation can potentially also have implications for how archaeological investigations should be directed at different altitudes in the region.

# Background
After the end of the Weichselian and the retreat of the Fennoscandian Ice Sheet, the isostatic rebound in south-eastern Norway has been so severe that the region has been subject to a continuous sea-level regression, meaning that the relative sea level has been dropping throughout prehistory. In other areas of Norway the situation is sometimes more complicated. In some areas relative sea-level change has transitioned between regression and transgression phases, and the archaeological remains can be destroyed. In other areas, the sea-level has been stable over longer periods of time, meaning coastal locations have often been reused multiple times over long time-spans, creating a palmpiset of settlement events that are impossible to disentangle. Comparatively, given the continuous and relatively rapid shoreline regression in south-eastern Norway, any one location in the region has only been shore-bound within a relatively limited time-span. As this general displacement trajectory is the same throughout the entirety of prehistory, this makes the region especially useful for testing of the assumption of a shore-bound settlement and to identify potential deviations from this pattern.

@@

# Methods
Relative sea level is so named because the parameters of interest is the height of the sea surface relative to its present level. Sea level is here understood as the distance from the solid surface of the Earth to the mean sea level as averaged over time, also known as the geoid. Variation in this distance is determined by effects that can be divided into three main categories. The first of these is eustatic sea-level change. Eustatic sea level is understood as the sea level if the water has been evenly distributed across the Earth's surface and without accounting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself. The eustatic sea-level is most notably impacted by glaciation and de-glaciation which can either bind or release large amounts of water into the oceans, as well as volumetric changes in the Earth's crust following from . The second category is that of istostasy, which pertains to adjustments in the Earth's crust that results from processes to regain gravitational equilibrium relative to the underlying viscous mantle. This is typically the result of erosion of the crust which causes its weight to be redistributed and the surface to be pushed upwards, or glacial istostasy which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, which either causes the lithosphere to be pushed into the mantle or to rebound and to lift upwards. Thirdly, relative sea-level is also impacted by thermostasy, which pertains to water temperatures and corresponding variation in water density.

@@

The goal is to test the synchroneity of two space-time processes, the occupation of a site and continuously shifting sea-levels. The synchroneity of these processes is tested by the comparison of singular events represented by the death of the dated organic materials. Additionally, defining what shore bound settlement represents can be problematic. In the literature this is often qualified with statements along the lines of "the sites were located on or close to the contemporenous shore-line". What constitutes close is of course highly variabl,e complicated further by the fact that vertical changes in sea level can results in widely different effects on the horisontal distance between a given point and the shore-line. Other elements of interest is the tidal range, and the exposure of the location in question, which can also have an impact on the distance of the site to the mean ocean surface.

>>>>

# Introduction
The association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C. Brøgger and his son, the archaeologist A.W. Brøgger, in their respective theses both published in 1905 (see also @maccurdy1906). Ever since, this finding has been fundamental to Stone Age research in Norway. One of its major implications is that relative sea-level change can be used for the dating of the sites. This is based on most Stone Age sites were situated on or close to the contemporary shoreline during their original occupation. A reconstruction of the trajectory of shoreline displacement can thus be used to date the original occupation of the sites based on their elevation relative to the present day sea level. The method of shoreline dating is still fundamental to Stone Age research in Norway, and is used both independently and to compliment the dating of sites where other sources of temporal data, such as typological indicators or radiometric dates, are either limited or not available. Given the generally coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating by means of shoreline displacement is often the only and most accurate method by which one can hope to date the sites. Dating by means of shoreline displacement is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because shoreline dating is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line. Consequently, adherence or deviation from this pattern has major implications for the socio-economic foundations of the societies in question.

Despite the fundamental role of shoreline-dating in Norway, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age. This can be used both to inform the certainty with which one can depend on dating by means of shoreline displacement, it can potentially identify societal shifts if there is chronological variation in the degree to which the sites are shore-bound, and any such variation can potentially also have implications for how archaeological investigations should be directed at different altitudes in the region. Furthermore, the method of Monte Carlo simulation to , can be extended to a wide range of use cases where one needs to evaluate and visualise the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.

# Background
Relative sea level is so named because the parameters of interest is the elevation of the sea surface relative to land. Relative sea level is the distance from the solid surface of the Earth to the mean sea level, as averaged over time, and is also known as the geoid. Variation in this relative distance follows from effects that can be divided into three main categories. The first of these is eustatic sea-level change. Eustatic sea-level is understood as the sea level if the water has been evenly distributed across the Earth's surface and without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself. The eustatic sea-level is mainly impacted by glaciation and de-glaciation, which can either bind or release large amounts of water into the oceans, as well as by volumetric changes in the Earth's crust following from . The second category is that of istostasy, which pertains to adjustments in the Earth's crust to regain gravitational equilibrium relative to the underlying viscous mantle. This is typically the result of glacial istostasy which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, and from erosion of the crust which causes its weight to be redistributed. These effects thus causes the lithosphere to be either pushed into the mantle due to increased weight or to rebound and to lift upwards due to a lower weight. Thirdly, relative sea-level is also impacted by thermostasy, which pertains to water temperatures and corresponding variation in water density.

Following the end of the Weichselian and the retreat of the Fennoscandian Ice Sheet, the isostatic rebound in south-eastern Norway has been so severe that the region has been subject to a continuous sea-level regression, despite corresponding eustatic sea-level rise. In other words, the relative sea level has been dropping throughout prehistory, a process that also contiues to this day. In other areas of Norway the situation is sometimes more complicated. On the outer coast, the relative sea-level change has transitioned between regression and transgression phases, and the archaeological remains can be destroyed. In other areas, the sea-level has been stable over longer periods of time, meaning coastal locations have often been reused multiple times over long time-spans, creating a palmpiset of settlement events that are impossible to disentangle. Comparatively, given the continuous and relatively rapid shoreline regression in south-eastern Norway, any one location in the region has only been shore-bound within a relatively limited time-span. As this general displacement trajectory is the same throughout the entirety of prehistory, this makes the region especially useful for testing of the assumption of a shore-bound settlement and to identify potential deviations from this pattern.

@@

@@

# Methods
The goal is effectively to test the synchroneity of two space-time processes: The occupation of a site and continuously shifting sea-levels. The synchroneity of these processes is tested by the comparison of singular events represented by the death of the dated organic materials. Additionally, defining what shore bound settlement represents can be problematic. In the literature this is often qualified with statements along the lines of "the sites were located on or close to the contemporenous shore-line". What constitutes close is of course highly variabl,e complicated further by the fact that vertical changes in sea level can results in widely different effects on the horisontal distance between a given point and the shore-line. Other elements of interest is the tidal range, and the exposure of the location in question, which can also have an impact on the distance of the site to the mean ocean surface.
